Bayesian regression modelling facilitates quantitative modelling of cell metabolism

### 2023-10-30 paper/main.qmd
Update paper

---
title: "Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism 
abstract: >
  This paper presents Maud, a command-line application that implements
  Bayesian statistical inference for kinetic models of biochemical metabolic
  reaction networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Below, we
  review the existing options in this area, explain how Maud improves on the state
  of the art, describe the intended modelling workflow and illustrate its use with
  an example application.
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
    - id: biosustain
      name: DTU
      department: DTU Biosustain
      city: Kongens Lyngby
      country: Denmark
  - name: "Nicholas Luke Cowie"
    affiliations: 
    - ref: biosustain
  - name: "Lars Keld Nielsen"
    affiliations: 
    - ref: biosustain
    - id: UQ
    - name: The University of Queensland
    - department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
    - city: St Lucia
    - country: Australia
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
      abbreviations: IR,NMR,UV
  acs-html:
   include-in-header: mathjax.html
   embed-resources: true
  docx: default
---

@@

### Computation

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and missing-
-data datasets. For the full dataset we obtained 1000 post-warmup samples each
from 4 independent Markov chains after 1000 warm-up samples and "hot-starting"
with a mass metric output by a previous model run.

For the missing-data dataset XXX.

For the Laplace approximation comparison, we used Maud’s Laplace mode. Full
details as well as instructions for reproducing our analysis can be found at
XXX.

@@

MCMC sampling for our comparison input yielded XXX samples in XXX time; Laplace
sampling yielded XXX samples in XXX time. The diagnostics indicated that our
algorithm was able to find the maximum a posteriori parameter configuration,
approximate the Hessian and use these quantities to generate approximate
posterior samples. The results can be found at <https://github.com/biosustain/Methionine_model/tree/main/results>.

@@

To further explore why this is the case we compared samples from the true
posterior and the Laplace approximation for the pairwise marginal distributions
of two Michaelis-Menten constants [XXX]- and [XXX]: see the bottom right cell of
figure @fig-laplace. This comparison demonstrates that the Laplace method is not
able to capture the correct relationships between parameters’ distributions.

@@

![](./figures/laplace.png)

CAPTION: FILL THIS IN!

@@

- Description of Maud's kinetic model including parameter dimensions and rate
equations.
- Description of the procedure used to generate the artificial case study
datasets.
- Specification of the prior distribution used for the case studies

>>>>

---
title: "Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism 
abstract: >
  This paper presents Maud, a command-line application that implements
  Bayesian statistical inference for kinetic models of biochemical metabolic
  reaction networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Below, we
  review the existing options in this area, explain how Maud improves on the state
  of the art, describe the intended modelling workflow and illustrate its use with
  an example application.
number-sections: true
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
    - id: biosustain
      name: DTU
      department: DTU Biosustain
      city: Kongens Lyngby
      country: Denmark
  - name: "Nicholas Luke Cowie"
    affiliations: 
    - ref: biosustain
  - name: "Lars Keld Nielsen"
    affiliations: 
    - ref: biosustain
    - id: UQ
    - name: The University of Queensland
    - department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
    - city: St Lucia
    - country: Australia
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
      abbreviations: IR,NMR,UV
#  acs-html:
#   include-in-header: mathjax.html
#   embed-resources: true
#  docx: default
---

@@

@@

MCMC sampling for our comparison input yielded 200 samples in 625 minutes;
Laplace sampling yielded the same number of samples in only one minute.
The diagnostics indicated that our algorithm was able to find the maximum
a posteriori parameter configuration, approximate the Hessian and use these
quantities to generate approximate posterior samples. The results can be found
at <https://github.com/biosustain/Methionine_model/tree/main/results>.

@@

To further explore why this is the case we compared samples from the true
posterior and the Laplace approximation for the pairwise marginal distributions
of two Michaelis-Menten constants $K_{m}^{A,r1}$ and $K_{cat}^{r1}$: see the
bottom right cell of figure @fig-laplace. This comparison demonstrates that
the Laplace method is not able to capture the correct relationships between
parameters’ distributions.

@@

![](../figures/laplace.png)

Graphical comparison of approximate posterior samples generated using Laplace
sampling (blue-green) with posterior samples generated using MCMC (dark
grey). (**Top left**) the two sets of samples clearly have different marginal
distributions for the overall log probability variable, indicating that the
Laplace samples do not accurately approximate the target distribution. (**Top
right**) The distribution of marginal posterior predictive log likelihood values
for training data flux measurements shows that the Laplace method tended to
yield much worse predictions compared with the true model. (**Bottom left**)
The Laplace method also tended to produce worse flux predictions for held-back
test measurements. (**Bottom right**) The marginal joint distribution of two
parameters: $K_m^{A,rl}$ and $K_{cat}^{r1}$. The Laplace method is not able
to track the correct joint distribution for this pair of parameters. This is
unsurprising given that the target distribution has position-dependent scales
which are difficult for a linear approximation to capture.

@@

- Description of Maud's kinetic model including parameter dimensions and rate
equations.
- Details of the procedure used to generate our case study results, including
generating the artificial case study datasets, specifying priors and carrying
out computation.
- Specification of the prior distribution used for the case studies

### 2023-10-30 paper/main.tex
Update paper

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\title[]{Bayesian regression modelling facilitates quantitative
modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{computation}{%
\subsubsection{Computation}\label{computation}}

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
missing- -data datasets. For the full dataset we obtained 1000
post-warmup samples each from 4 independent Markov chains after 1000
warm-up samples and ``hot-starting'' with a mass metric output by a
previous model run.

For the missing-data dataset XXX.

For the Laplace approximation comparison, we used Maud's Laplace mode.
Full details as well as instructions for reproducing our analysis can be
found at XXX.

@@

MCMC sampling for our comparison input yielded XXX samples in XXX time;
Laplace sampling yielded XXX samples in XXX time. The diagnostics
indicated that our algorithm was able to find the maximum a posteriori
parameter configuration, approximate the Hessian and use these
quantities to generate approximate posterior samples. The results can be
found at
\url{https://github.com/biosustain/Methionine_model/tree/main/results}.

@@

To further explore why this is the case we compared samples from the
true posterior and the Laplace approximation for the pairwise marginal
distributions of two Michaelis-Menten constants {[}XXX{]}- and
{[}XXX{]}: see the bottom right cell of figure Figure~\ref{fig-laplace}.
This comparison demonstrates that the Laplace method is not able to
capture the correct relationships between parameters' distributions.

@@

\includegraphics{./figures/laplace.png}

@@

\caption{\label{fig-laplace}CAPTION: FILL THIS IN!}

@@

\begin{itemize}
\tightlist
\item
  Description of Maud's kinetic model including parameter dimensions and
  rate equations.
\item
  Description of the procedure used to generate the artificial case
  study datasets.
\item
  Specification of the prior distribution used for the case studies
\end{itemize}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\title[]{Bayesian regression modelling facilitates quantitative
modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi

@@

@@

MCMC sampling for our comparison input yielded 200 samples in 625
minutes; Laplace sampling yielded the same number of samples in only one
minute. The diagnostics indicated that our algorithm was able to find
the maximum a posteriori parameter configuration, approximate the
Hessian and use these quantities to generate approximate posterior
samples. The results can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/results}.

@@

To further explore why this is the case we compared samples from the
true posterior and the Laplace approximation for the pairwise marginal
distributions of two Michaelis-Menten constants \(K_{m}^{A,r1}\) and
\(K_{cat}^{r1}\): see the bottom right cell of figure
Figure~\ref{fig-laplace}. This comparison demonstrates that the Laplace
method is not able to capture the correct relationships between
parameters' distributions.

@@

\includegraphics{../figures/laplace.png}

@@

\caption{\label{fig-laplace}Graphical comparison of approximate
posterior samples generated using Laplace sampling (blue-green) with
posterior samples generated using MCMC (dark grey). (\textbf{Top left})
the two sets of samples clearly have different marginal distributions
for the overall log probability variable, indicating that the Laplace
samples do not accurately approximate the target distribution.
(\textbf{Top right}) The distribution of marginal posterior predictive
log likelihood values for training data flux measurements shows that the
Laplace method tended to yield much worse predictions compared with the
true model. (\textbf{Bottom left}) The Laplace method also tended to
produce worse flux predictions for held-back test measurements.
(\textbf{Bottom right}) The marginal joint distribution of two
parameters: \(K_m^{A,rl}\) and \(K_{cat}^{r1}\). The Laplace method is
not able to track the correct joint distribution for this pair of
parameters. This is unsurprising given that the target distribution has
position-dependent scales which are difficult for a linear approximation
to capture.}

@@

\begin{itemize}
\tightlist
\item
  Description of Maud's kinetic model including parameter dimensions and
  rate equations.
\item
  Details of the procedure used to generate our case study results,
  including generating the artificial case study datasets, specifying
  priors and carrying out computation.
\item
  Specification of the prior distribution used for the case studies
\end{itemize}
